Hot and fresh: Pervasive climate stressors of seagrass in a large Gulf coast estuary

### 2024-01-23 manu-draft.qmd
text edits and minor comments in line with text

Climate change is an existential threat to natural resources that sustain human health and the environment. In Tampa Bay, Florida (USA), seagrass coverage has declined since 2016 by nearly a third (11,518 acres), despite sustained reductions of nitrogen loads supportive of light environments for growth. Changing physical water quality conditions related to climate change may be stressing seagrasses beyond their optimal growth ranges, requiring an assessment to determine if this decline can be linked to climate stress. Three datasets of varying sample design and coverage were evaluated to characterize physical environments in Tampa Bay and the potential relationships with seagrass change. Over the last fifty years, Tampa Bay has become hotter and fresher with water temperature increasing by 0.03 - 0.07 $^\circ$C per year and salinity decreasing by 0.04 - 0.2 PSU per year. These changes varied spatially and seasonally, with the most dramatic changes observed in the upper bay segments during the summer months. Simple linear models provided limited evidence that recent seagrass declines are associated with hotter and fresher conditions. Trends in warming and increased precipitation in the region are likely to continue, further creating suboptimal conditions for seagrasses in Tampa Bay. These results should compel resource managers to consider the likelihood that reduced resilience of estuarine resources due to shifting ecological baselines from climate change could complicate long-standing management paradigms. While conventional management approaches that focus on limiting nutrient loads should be continued, their future effectiveness may be limited and warrant additional, complementary interventions to maintain ecosystem health into the future.

@@

The monitoring and management of seagrasses in coastal environments has received substantial attention at a global scale. Seagrasses are fundamental indicators of ecosystem health [@roca2016; @orth2017], while also serving as foundational species that provide numerous ecosystem services [@fourqurean2012; @orth2020; @orth2023]. Seagrasses have been in global decline with rapid development of coastal environments, particularly in the latter half of the 20th century with accelerating losses estimated at a rate of 110 km$^2$ yr$^{-1}$ since the 1980s [@waycott2009; @dunic2021]. These losses are comparable to, if not more significant than, other critical coastal environments [@duarte2008a], which threatens the natural and economic services that these habitats provide. Losses have been attributed to numerous stressors including decline in light environments with nutrient enrichment, sedimentation, and physical disturbance primarily from human activities [@duarte1995; @orth2006; @burkholder2007]. Furthermore, natural disturbances such as storm events and disease have also been implicated [@robblee1991; @tomasko2020]. Contemporary management actions and interventions aimed at mitigating loss and ultimately supporting restoration require adaptive approaches to address the effects of multiple stressors that have contributed to seagrass decline [@dunic2023].

@@

Climate change has complicated the understanding of ecosystem response to conventional stressors, presenting new challenges and expectations for how ecological resources will respond to management actions [@statham2012a; @sherwood2013]. In addition to sea-level rise, the most anticipated effects of climate change in coastal environments are increasing temperature and alteration of precipitation patterns. These changes will profoundly alter physical habitats, creating suboptimal or uninhabitable conditions for many species [@madeira2012; @lefcheck2017; @hensel2023]. @lefcheck2017 evaluated interactive effects of water clarity and rising temperatures on seagrasses with over 30 years of data in Chesapeake Bay. The effects of decreased water clarity for seagrasses in shallow habitats were exacerbated by acute warming related to climate change, demonstrating a concerning synergy of multiple stressors. Species shifts are also expected to occur as changing physical conditions decrease the competitive advantages of historically abundant species. Again in Chesapeake Bay, the abundance of widgeongrass (*Ruppia maritima*) has been mediated by nutrient reductions, while also outcompeting and replacing formerly dominant eelgrass (*Zostera marina*) given higher tolerance to increased temperatures [@hensel2023]. Changing frequency and severity of precipitation patterns may further alter the distribution and abundance of seagrasses by pushing species outside of optimal salinity tolerance ranges [@rasheed2011; @webster2021]. These changes may produce hypersaline conditions under periods of prolonged drought or more freshwater conditions with increased storm events. Seagrasses may respond differently under individual or multiple stressors [@lefcheck2017; @zhang2023], suggesting place-based assessments of response to climate change provide valuable insights for managers responsible for maintaining ecological resources.

In Tampa Bay, Florida, seagrasses are a primary indicator of bay health that have been the focus of management efforts over the last thirty years [@sherwood2017]. Through successful reduction of nitrogen loads by 2/3 pre-regulation, seagrasses have recovered from a low in the 1980s to an all-time high in 2016 of 41,655 acres [16,857 ha, @greening2006; @greening2014]. Seagrass distribution and abundance has been dynamic throughout this period, overall responding positively to increases in water clarity with nutrient load reductions, while also demonstrating more short-term variability in response to regional climate events [@greening2006]. Since 2016, seagrass areal cover has decreased by nearly 1/3 despite relatively stable water quality with adequate light penetration that has been historically shown to support growth [@tbep0696; @Beck20a]. Factors that have influenced this recent decline are unknown and the effects of climate change on physical conditions have been implicated as potential drivers. Following global trends, recent work has demonstrated a broad long-term trend of increasing water temperature in Tampa Bay, although at a relatively coarse scale [@nickerson2023]. Changing salinity has not been well-described, nor have potential links of changing salinity and temperature with recent seagrass change. Tampa Bay is rich with historical data that can be used to evaluate long-term trends and to develop models that can explain relationships of seagrass change with water quality conditions. This information can fill a critical knowledge gap that can inform regional management activities, while also demonstrating the confounding effects of climate change with ecosystem response to conventional stressors.

This paper describes a comprehensive assessment of long-term trends in water temperature and salinity in the Tampa Bay estuary over the last fifty years. Three datasets of varying sample design and temporal coverage were used to assess the primary hypothesis that Tampa Bay is trending towards hotter and fresher conditions that are likely pushing seagrasses beyond their optimal tolerance ranges. This analysis was supported by 30-year seagrass datasets describing approximate biennial coverage of all seagrasses, annual transect monitoring describing cover of individual species, and synoptic seagrass data collected with the water quality surveys. Specific focus was on the periods prior to and after 2016, when seagrasses recovered baywide and then declined despite water quality conditions remaining relatively stable. Our primary goal was to relate recent seagrass changes to climate stressors to provide a motivating example for the management community in the Tampa Bay region that a shifting ecological baseline presents new challenges beyond the known success story of long-term bay recovery. The results are also provided as a cautionary example of the subtle but increasing role of climate-related stressors on the resiliency of estuarine systems at the tropical-temperate boundary, prompting a reassessment of existing management paradigms to accommodate current and anticipated future conditions.

@@

Tampa Bay is the largest open-water estuary in Florida covering 400 mi$^2$ (1,036 km$^2$) and the second largest in the Gulf of Mexico. The watershed covers an additional 2,200 mi$^2$ (5,872 km$^2$) with the Hillsborough, Alafia, Manatee, and Little Manatee rivers contributing a majority of freshwater inflow to the bay. Tampa Bay straddles the tropical and temperate boundary of central Florida characterized by warm, humid conditions and a distinct rainy season during the summer. The watershed is heavily developed and includes over 3 million people with 42% of the land as urban or suburban contributing substantial inputs of wastewater and stormwater runoff that can stress bay resources [@beck2023]. The geology of the watershed is rich in phosphates and mining activities have greatly altered the landscape, with notable spills and releases of wastewater that have affected water quality and biological resources [@garrett2011; @beck2022a]. Important subtidal habitats include seagrasses, hard bottom, tidal flats, and oyster reefs, where a majority of management efforts have historically focused on restoring and maintaining seagrass cover [@sherwood2017]. Additional native habitats include intertidal wetlands (mangroves, salt marshes, salt barrens) and pine forests, oak hammocks, and freshwater wetlands present in upland habitats [@Robison20]. Losses of native uplands and potentially restorable habitats to development in the watershed from 1990 to 2020 have been estimated at 188,429 acres [76,254 ha, @beck2023].

Tampa Bay is divided into distinct sub-segments defined by physical and natural boundaries to assist with water quality management activities [@lewis1985]: Old Tampa Bay (OTB) in the northwest; Hillsborough Bay (HB) in the northeast; Middle Tampa Bay (MTB); and Lower Tampa Bay (LTB) that connects to the Gulf of Mexico (Figure [-@fig-map]a). Old Tampa Bay and Hillsborough Bay have historically had the most degraded water quality primarily from nutrient inputs from wastewater and stormwater [@greening2014]. Hydrologic conditions vary between the two, such that Hillsborough Bay receives a majority of direct surface water inflow from the Hillsborough and Alafia rivers, whereas Old Tampa Bay receives much less inflow with a majority from multiple small tributaries and manmade flood control conveyances [@tbep0623]. Notably, Old Tampa Bay has restricted circulation from multiple land bridges that traverse the bay [@sherwood2015; @luther2022]. Recurring seasonal harmful algal blooms of the dinoflagellete *Pyrodinium bahamense* have contributed to exceedances of the chlorophyll-a regulatory standard in Old Tampa Bay [@lopez2023]. By comparison, water quality conditions in Middle Tampa Bay and Lower Tampa Bay are generally better than the upper two bay segments primarily from more frequent water exchanges with the Gulf of Mexico and lower nutrient loading [@tbep0623]. All bay segments are relatively shallow, with a baywide mean depth of approximately 3 m. Light penetration typically reaches bottom habitats under current conditions, although seagrasses were historically limited by high phytoplankton production that affected light environments prior to wastewater regulation, particularly in Old Tampa Bay and Hillsborough Bay [@greening2014; @johansson15].

@@

Several datasets with distinct sample designs are available to assess long-term changes in water temperature and salinity in Tampa Bay. These datasets were evaluated individually to assess trends and relationships with seagrass change to provide a weight-of-evidence approach for potential causal relationships driving the recent decline. First, the Environmental Protection Commission (EPC) of Hillsborough County has collected discrete water quality measurements monthly at 45 stations in the major bay segments since the late 1970s (Figure [-@fig-map]b). These data are used for regulatory assessments for the nutrient TMDL <!--# spell out --> in Tampa Bay. Water quality samples are collected at each station from surface water grabs (e.g., nutrients) or *in situ* measurements of physical parameters (e.g., salinity, temperature) collected at the surface, mid-depth, and bottom. Time of sampling can vary, although most samples are collected from mid-morning to early afternoon. All surface and bottom salinity and temperature measurements for each of the 45 monitoring stations were evaluated herein. Trends were assessed for both surface and bottom samples, as described below, whereas only the bottom measurements were used for comparison to seagrass trends. The data were obtained using the tbeptools R package that imports the data directly from a stable web address provided by the EPC [@beck2021].

@@

The third and final dataset evaluated was from the Pinellas County Department of Environmental Management (PDEM). Surface waters in Pinellas County have been monitored since the 1990s, although a consistent stratified random sampling designed has been used in Tampa Bay since 2003 primarily to support robust statistical assessments for NPDES <!--# spell out --> reporting. Data were obtained by request to PDEM staff for the western portion of Old Tampa Bay where sampling occurs from 2003 to present (Figure [-@fig-map]d, also available at <https://wateratlas.usf.edu/>). We focused primarily on OTB for the analysis of the PDEM data given the length of record, consistency of sampling, and relative loss of seagrass compared to the other bay segments. Four distinct spatial zones in OTB are used to stratify the random selection of sample points that typically include 4 sample points per month in each zone. Water quality samples at each site are similar to those collected by EPC, where only bottom measurements for salinity and temperature were retained for analysis. Seagrass presence/absence is also recorded at each site and all sites were defined as "seagrass" if only seagrass species were identified (any with macroalgae were excluded) and "no seagrass" if bare sediment was observed.

@@

The first goal of the analysis was to describe spatial and temporal trends in water temperature and salinity using the three water quality datasets described above. This assessment provided an indication of the extent of change in Tampa Bay as context for understanding potential relationships with seagrass change. An assumption was that any changes in physical characteristics in Tampa Bay were driven by interannual changes in weather conditions related to long-term (multi-decadal) climate change. For comparison to water quality conditions, daily air temperature data were obtained from Tampa International Airport [using the *rnoaa* R package, @chamberlain23] and monthly rainfall data for the Tampa Bay watershed were obtained from the SWFWMD (<https://www.swfwmd.state.fl.us/resources/data-maps/rainfall-summary-data-region>). Only rainfall data for the rainy season (June to September) were evaluated. Additionally, the Standardized Precipitation Index [SPI, @begueria2013] was estimated from the rainfall data to identify periods of time when rainfall significantly deviated from the long-term average [using the *spei* R package, @begueria23]. Annual hydrologic loading data to Tampa Bay beginning in 1985 were also obtained for comparison to annual precipitation [@tbep0623]. All climate and loading data were evaluated annually with simple linear regression trends to assess change over time. Water temperature and salinity trends using the EPC, FIM, and PDEM data were similarly evaluated by averaging the monthly data each year for each bay segment. Linear trends for these data were evaluated based on averages of all stations and months within each year and bay segment, which allowed for comparable statistical power between datasets with different sample designs.

@@

A fundamental challenge describing niche space is identifying the boundaries for optimal conditions. In Tampa Bay, the three dominant seagrass species are *Halodule wrightii* (shoal grass), *Syringodium filiforme* (manatee grass), and *Thalassia testudinum* (turtle grass) [@lewis1985]. These species co-occur often in mixed beds throughout the bay, although some differences in abundance are observed across salinity ranges. Shoal grass is tolerant of a wide range of salinity [@lirman2003], but is more abundant in oligo/mesohaline portions of Tampa Bay. Conversely, turtle grass is less tolerant of low salinity and is more abundant in more euryhaline conditions near the mouth of Tampa Bay. Reported salinity ranges for each of these species varies depending on location, season, and other co-occurring factors like temperature [@phillips1960; @mcmillan1967; @zieman1975; @lewis1985], although most studies place lower limits of salinity in the range of 15-25 ppt. Optimal temperature ranges are similar between species, with reduced growth observed at temperatures above 30$^\circ$ C [@zieman1975; @lewis1985].

Because of the uncertainty in defining thresholds for optimal temperature and salinity ranges, multiple thresholds were evaluated to describe the potential for stress and how it may be related to changes in seagrass. Distinctions were not made between species, primarily due to lack of consensus between studies and likely site-specific ranges that affect seagrass growth in Tampa Bay. First, we developed metrics of potential temperature and salinity stress by quantifying the maximum number of continuous days each year when temperature was above or salinity was below a given threshold. This approach assumed that stress could be observed based on duration of exposure (i.e., maximum number of continuous days each year) relative to a threshold that may are may not be outside of the optimum range for seagrasses. These metrics were quantified from the monthly long-term observations in the EPC data. To quantify daily counts each year, a continuous prediction of temperature and salinity over time at each of 45 stations was estimated using Generalized Additive Models (GAMs) fit to temperature or salinity with a single predictor for decimal year [@wood17]. The smoothing spline for decimal year had sufficient knots to capture the seasonal signal within each year and the long-term inter-annual trends. Model fit for each station was sufficient to calculate daily predictions to assess potential stressor metrics (Figure S5, R$^2$ ranged from `r fitest$temp$minr2` to `r fitest$temp$maxr2` for temperature models, `r fitest$sal$minr2` to `r fitest$sal$maxr2` for salinity models, Tables S1, S2).

@@

For comparison to seagrass, the annual metrics were referenced to approximate periods of time between the annual seagrass transect surveys, as opposed to the calendar year for describing trends. Transect surveys are typically done in the late summer during peak growing season. The average sampling date for each year across the subset of transects in each bay segment was estimated and the stressor metrics were referenced accordingly. For example, if the average transect date for a bay segment was September 15th in 2009 and Oct 1st in 2010, the 2010 stressor counts were based on all days between the two average dates. This was repeated for all years from 1998 to 2022 when transect data were available. Bay segment stressor metrics were calculated as the average counts in each "transect year" from all stations in each segment. The stressor metrics were compared to frequency occurrence (all species) each year by bay segment. Generalized Linear regression Models (GLMs) were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, both, bay segment, and time period (recovery pre-2016, decline post-2016). Two models were evaluated, one with the temperature and salinity metrics together and another with the both metric given that the latter was a combined metric of the former. Interactions were included between temperature, salinity, and time period or the both metric and time period, depending on the model. Bay segment was included as a main effect without interactions. The most parsimonious models were identified by forward and backward selection of the independent variables and comparison with AIC values [@sakamoto1986; @venables2002b]. All models excluded Lower Tampa Bay because of minimal seagrass change over time.

@@

Long-term meteorological data showed increasing trends for air temperature and precipitation ([@fig-meteowqraw]). Mean annual air temperature has increased by `r airtempslo` $^\circ$C per year (`r airtemppvl`, $R^2$ `r airtemprsq`). Mean annual air temperature in 1975 was `r airtempstr` $^\circ$C, whereas current mean annual air temperature in 2022 was `r airtempend` $^\circ$C, showing an overall increase in the period of record of `r airtempdif` $^\circ$C. Similarly, total precipitation during the rainy season has increased by `r prpslo` mm per year, although the model was not significant (`r prppvl`, $R^2$ `r prprsq`). Removing September from the rainy season showed a significantly increasing trend of `r prpslo2` mm per year (`r prppvl2`, $R^2$ `r prprsq2`). Using this model, total precipitation in 1975 was `r prpstr2` mm, whereas current total precipitation in 2022 was `r prpend2` mm, showing an overall increase in the period of record of `r prpdif2` mm. Trends in total annual hydrologic load from 1985 to present followed total annual precipitation, although the linear model was not significant likely due to a shorter period of record. The SPI showed notable anomalies in precipitation, with pronounced rainy periods in the early 1980s, late 1990s, 2005, and 2015-2020.

>>>>

Climate change is an existential threat to natural resources that sustain human health and the environment. In Tampa Bay, Florida (USA), seagrass coverage has declined since 2016 by nearly a third (11,518 acres), despite sustained reductions of nitrogen loads supportive of light environments for growth. Changing physical water quality conditions related to climate change may be stressing seagrasses beyond their optimal growth ranges, requiring an assessment to determine if this decline can be linked to climate stress. Three datasets of varying sampling design and coverage were evaluated to characterize physicochemical environments in Tampa Bay and the potential relationships with seagrass change. Over the last fifty years, Tampa Bay has become hotter and fresher with water temperature increasing by 0.03 - 0.07 $^\circ$C per year and salinity decreasing by 0.04 - 0.2 PSU per year. These changes varied spatially and seasonally, with the most dramatic changes observed in the upper bay during the summer months. Simple linear models provided limited evidence that recent seagrass declines are associated with hotter and fresher conditions. Trends in warming and increased precipitation in the region are likely to continue, further creating suboptimal conditions for seagrasses in Tampa Bay. These results should compel resource managers to consider the likelihood that reduced resilience of estuarine resources due to shifting ecological baselines from climate change could complicate long-standing management paradigms. While conventional management approaches that focus on limiting nutrient loads should be continued, their future effectiveness may be limited and warrant additional, complementary interventions to maintain ecosystem health into the future.

@@

The monitoring and management of seagrasses in coastal environments has received substantial attention at a global scale. Seagrasses are fundamental indicators of ecosystem health [@roca2016; @orth2017], while also serving as foundational species that provide numerous ecosystem services [@fourqurean2012; @orth2020; @orth2023]. Seagrasses have been in global decline with rapid development of coastal environments, particularly in the latter half of the 20th century with accelerating losses estimated at a rate of 110 km$^2$ yr$^{-1}$ since the 1980s [@waycott2009; @dunic2021]. These losses are comparable to, if not more significant than, other interconnected critical coastal environments such as mangroves and salt marshes [@duarte2008a <!--# any more recent references? -->], which threatens the natural and economic services that these habitats provide. Losses have been attributed to numerous stressors including decline in light environments with nutrient enrichment, sedimentation, and physical disturbance primarily from human activities [@duarte1995; <!--# Add Hall et al. 1999, Florida bay seagrass dieoff and recovery --> @orth2006; @burkholder2007]. Furthermore, natural disturbances such as storm events and disease have also been implicated [@robblee1991; @tomasko2020]. Contemporary management actions and interventions aimed at mitigating loss and ultimately supporting restoration require adaptive approaches to address the effects of multiple stressors that have contributed to seagrass decline [@dunic2023].

@@

Climate change has complicated the understanding of ecosystem response to conventional stressors, presenting new challenges and expectations for how ecological resources will respond to management actions [@statham2012a; @sherwood2013]. In addition to sea-level rise, the most anticipated effects of climate change in coastal environments are increasing temperature and alteration of precipitation patterns. These changes will profoundly alter physicochemical habitats, creating suboptimal or uninhabitable conditions for many species [@madeira2012; @lefcheck2017; @hensel2023]. @lefcheck2017 evaluated interactive effects of water clarity and rising temperatures on seagrasses with over 30 years of data in Chesapeake Bay. The effects of decreased water clarity for seagrasses in shallow habitats were exacerbated by acute warming related to climate change, demonstrating a concerning synergy of multiple stressors. Species shifts are also expected to occur as changing physical conditions decrease the competitive advantages of historically abundant species. Again in Chesapeake Bay, the abundance of widgeongrass (*Ruppia maritima*) has been mediated by nutrient reductions, while also outcompeting and replacing formerly dominant eelgrass (*Zostera marina*) given higher tolerance to increased temperatures [@hensel2023]. Changing frequency and severity of precipitation patterns may further alter the distribution and abundance of seagrasses by pushing species outside of optimal salinity tolerance ranges [@rasheed2011; @webster2021]. These changes may produce hypersaline conditions under periods of prolonged drought or more freshwater conditions with increased storm events. Seagrasses may respond differently under individual or multiple stressors [@lefcheck2017; @zhang2023], suggesting place-based assessments of response to climate change provide valuable insights for managers responsible for maintaining ecological resources.

In Tampa Bay, Florida, seagrasses are a primary indicator of bay health that have been the focus of management efforts over the last thirty years [@sherwood2017]. Through successful reduction of nitrogen loads, seagrasses have recovered from a low in the 1980s to an all-time high in 2016 of 41,655 acres [16,857 ha, @greening2006; @greening2014]. Seagrass distribution and abundance has been dynamic throughout this period, overall responding positively to increases in water clarity with nutrient load reductions, while also demonstrating more short-term variability in response to regional climate events [@greening2006]. Since 2016, seagrass areal cover has decreased by nearly 1/3 despite relatively stable water quality with adequate light penetration that has been historically shown to support growth [@tbep0696; @Beck20a]. Factors that have influenced this recent decline are unknown and the effects of climate change on physical conditions have been implicated as potential drivers. Following global trends, recent work has demonstrated a broad long-term trend of increasing water temperature in Tampa Bay, although at a relatively coarse scale [@nickerson2023]. Changing salinity has not been well-described, nor have potential links of changing salinity and temperature with recent seagrass change. Tampa Bay is rich with historical data that can be used to evaluate long-term trends and to develop models that can explain relationships of seagrass change with water quality conditions. This information can fill a critical knowledge gap that can inform regional management activities, while also demonstrating the confounding effects of climate change with ecosystem response to conventional stressors.

This paper describes a comprehensive assessment of long-term trends in water temperature and salinity in the Tampa Bay estuary over the last fifty years. Three datasets of varying sampling designs and temporal coverage were used to assess the primary hypothesis that Tampa Bay is trending towards hotter and fresher conditions that are likely pushing seagrasses beyond their optimal tolerance ranges. This analysis was supported by 30-year seagrass datasets describing approximate biennial coverage of all seagrasses, annual transect monitoring describing cover of individual species, and synoptic seagrass data collected with the water quality surveys. Specific focus was on the periods prior to and after 2016, when seagrasses recovered baywide and then declined despite water quality conditions remaining relatively stable. Our primary goal was to relate recent seagrass changes to climate stressors to provide a motivating example for the management community in the Tampa Bay region that a shifting ecological baseline presents new challenges beyond the known success story of long-term bay recovery. The results are also provided as a cautionary example of the subtle but increasing role of climate-related stressors on the resiliency of estuarine systems at the tropical-temperate boundary, prompting a reassessment of existing management paradigms to accommodate current and anticipated future conditions.

@@

Tampa Bay is the largest open-water estuary in Florida covering 400 mi$^2$ (1,036 km$^2$) and the second largest in the Gulf of Mexico. The watershed covers an additional 2,200 mi$^2$ (5,872 km$^2$) with the Hillsborough, Alafia, Manatee, and Little Manatee rivers that contribute to a majority of freshwater inflow to the bay. Tampa Bay straddles the tropical and temperate boundary of central Florida characterized by warm, humid conditions and a distinct rainy season during the summer<!--# reference? -->. The watershed is heavily developed and includes over 3 million people <!--# reference economic valuation? -->with 42% of the land as urban or suburban contributing substantial inputs of wastewater and stormwater runoff that can stress bay resources [@beck2023]. The geology of the watershed is rich in phosphates and mining activities have greatly altered the landscape, with notable spills and releases of wastewater that have affected water quality and biological resources [@garrett2011; @beck2022a]. Important subtidal habitats include seagrasses, hard bottom, tidal flats, and oyster reefs, where a majority of management efforts have historically focused on restoring and maintaining seagrass cover [@sherwood2017]. Additional native habitats include intertidal wetlands (mangroves, salt marshes, salt barrens) and pine forests, oak hammocks, and freshwater wetlands present in upland habitats [@Robison20]. Losses of native uplands and potentially restorable habitats to development in the watershed from 1990 to 2020 have been estimated at 188,429 acres [76,254 ha, @beck2023].

Tampa Bay is divided into distinct sub-segments defined by physical and natural boundaries to assist with water quality management activities [@lewis1985]: Old Tampa Bay (OTB) in the northwest; Hillsborough Bay (HB) in the northeast; Middle Tampa Bay (MTB); and Lower Tampa Bay (LTB) that connects to the Gulf of Mexico (Figure [-@fig-map]a). Old Tampa Bay and Hillsborough Bay have historically had the most degraded water quality primarily from nutrient inputs from wastewater and stormwater [@greening2014]. Hydrologic conditions vary between the two, such that Hillsborough Bay receives a majority of direct surface water inflow from the Hillsborough and Alafia rivers, whereas Old Tampa Bay receives much less inflow with a majority from multiple small tributaries and manmade flood control conveyances [@tbep0623]. Notably, Old Tampa Bay has restricted circulation from multiple land bridges <!--# use causeways instead? -->that traverse the bay [@sherwood2015; @luther2022]. Recurring seasonal harmful algal blooms of the dinoflagellete *Pyrodinium bahamense* have contributed to exceedances of the chlorophyll-a regulatory standard in Old Tampa Bay [@lopez2023]. By comparison, water quality conditions in Middle Tampa Bay and Lower Tampa Bay are generally better than the upper two bay segments primarily from more frequent water exchanges with the Gulf of Mexico and lower nutrient loading [@tbep0623]. All bay segments are relatively shallow, with a baywide mean depth of approximately 3 m. Light penetration typically reaches bottom habitats under current conditions, although seagrasses were historically limited by high phytoplankton production that affected light environments prior to wastewater regulation, particularly in Old Tampa Bay and Hillsborough Bay [@greening2014; @johansson15].

@@

Several datasets with distinct sample designs are available to assess long-term changes in water temperature and salinity in Tampa Bay. These datasets were evaluated individually to assess trends and relationships with seagrass change to provide a weight-of-evidence approach for potential causal relationships driving the recent decline. First, the Environmental Protection Commission (EPC) of Hillsborough County has collected discrete water quality measurements monthly at 45 stations in the major bay segments since the late 1970s (Figure [-@fig-map]b). These data are used for regulatory assessments for the nutrient Total Maximum Daily Load in Tampa Bay. Water quality samples are collected at each station from surface water grabs (e.g., nutrients) or *in situ* measurements of physical parameters (e.g., salinity, temperature) collected at the surface, mid-depth, and bottom. Time of sampling can vary, although most samples are collected from mid-morning to early afternoon. All surface and bottom salinity and temperature measurements for each of the 45 monitoring stations were evaluated herein. Trends were assessed for both surface and bottom samples, as described below, whereas only the bottom measurements were used for comparison to seagrass trends. The data were obtained using the tbeptools R package that imports the data directly from a stable web address provided by the EPC [@beck2021].

@@

The third and final dataset evaluated was from the Pinellas County Department of Environmental Management (PDEM). Surface waters in Pinellas County have been monitored since the 1990s, although a consistent stratified random sampling designed has only been used in Tampa Bay since 2003 primarily to support robust statistical assessments for NPDES <!--# spell out --> reporting. Data were obtained by request to PDEM staff for the western portion of Old Tampa Bay where sampling occurs from 2003 to present (Figure [-@fig-map]d, also available at <https://wateratlas.usf.edu/>). We focused primarily on OTB for the analysis of the PDEM data given the length of record, consistency of sampling, and relative loss of seagrass compared to the other bay segments. Four distinct spatial zones in OTB are used to stratify the random selection of sample points that typically include 4 sample points per month in each zone. Water quality samples at each site are similar to those collected by EPC, where only bottom measurements for salinity and temperature were retained for analysis. Seagrass presence/absence is also recorded at each site and all sites were defined as "seagrass" if only seagrass species were identified (any with macroalgae were excluded) and "no seagrass" if bare sediment was observed.

@@

The first goal of the analysis was to describe spatial and temporal trends in water temperature and salinity using the three water quality datasets described above. This assessment provided an indication of the extent of change in Tampa Bay as context for understanding potential relationships with seagrass change. An assumption was that any changes in physical characteristics in Tampa Bay were driven by interannual changes in weather conditions related to long-term (multi-decadal) climate change. For comparison to water quality conditions, daily air temperature data were obtained from Tampa International Airport [using the *rnoaa* R package, @chamberlain23] and monthly rainfall data for the Tampa Bay watershed were obtained from the SWFWMD (<https://www.swfwmd.state.fl.us/resources/data-maps/rainfall-summary-data-region>). Only rainfall data for the rainy season (June to September) were evaluated. Additionally, the Standardized Precipitation Index [SPI, @begueria2013] was estimated from the rainfall data to identify periods of time when rainfall significantly deviated from the long-term average [using the *spei* R package, @begueria23]. Annual hydrologic loading data to Tampa Bay beginning in 1985 were also obtained for comparison to annual precipitation [@tbep0623]. All climate and loading data were evaluated annually with simple linear regression trends to assess change over time. Water temperature and salinity trends using the EPC, FIM, and PDEM data were similarly evaluated by averaging the monthly data each year for each bay segment. Linear trends for these data were evaluated based on averages of all stations and months within each year and bay segment, which allowed for comparable statistical power between datasets with different sampling designs.

@@

A fundamental challenge describing niche space is identifying the boundaries for optimal conditions. In Tampa Bay, the three dominant seagrass species are *Halodule wrightii* (shoal grass), *Syringodium filiforme* (manatee grass), and *Thalassia testudinum* (turtle grass) [@lewis1985]. These species co-occur often in mixed beds throughout the bay, although some differences in abundance are observed across salinity ranges. Shoal grass is tolerant of a wide range of salinity [@lirman2003], but is more abundant in oligo/mesohaline portions of Tampa Bay. Conversely, turtle grass is less tolerant of low salinity and is more abundant in more euryhaline conditions near the mouth of Tampa Bay. Reported salinity ranges for each of these species varies depending on location, season, and other co-occurring factors like temperature [@phillips1960; @mcmillan1967; @zieman1975; @lewis1985], although most studies place lower limits of salinity in the range of 15-25 ppt. Optimal temperature ranges are similar between species, with reduced growth observed at temperatures above 30$^\circ$ C [@zieman1975; @lewis1985]. <!--# What about Syringodium? Found in deeper water, any info on temperature range -->

Because of the uncertainty in defining thresholds for optimal temperature and salinity ranges, multiple thresholds were evaluated to describe the potential for stress and how it may be related to changes in seagrass. Distinctions were not made between species, primarily due to lack of consensus between studies and likely site-specific ranges that affect seagrass growth in Tampa Bay. First, we developed metrics of potential temperature and salinity stress by quantifying the maximum number of continuous days each year when temperature was above or salinity was below a given threshold. This approach assumed that stress could be observed based on duration of exposure (i.e., maximum number of continuous days each year) relative to a threshold that may or may not be outside of the optimum range for seagrasses. These metrics were quantified from the monthly long-term observations in the EPC data. To quantify daily counts each year, a continuous prediction of temperature and salinity over time at each of 45 stations was estimated using Generalized Additive Models (GAMs) fit to temperature or salinity with a single predictor for decimal year [@wood17]. The smoothing spline for decimal year had sufficient knots to capture the seasonal signal within each year and the long-term inter-annual trends. Model fit for each station was sufficient to calculate daily predictions to assess potential stressor metrics (Figure S5, R$^2$ ranged from `r fitest$temp$minr2` to `r fitest$temp$maxr2` for temperature models, `r fitest$sal$minr2` to `r fitest$sal$maxr2` for salinity models, Tables S1, S2).

@@

For comparison to seagrass, the annual metrics calculated from the EPC data were referenced to approximate periods of time between the annual seagrass transect surveys, as opposed to the calendar year for describing trends. Transect surveys are typically done in the late summer during peak growing season. The average sampling date for each year across the subset of transects in each bay segment was estimated and the stressor metrics were referenced accordingly. For example, if the average transect date for a bay segment was September 15th in 2009 and Oct 1st in 2010, the 2010 stressor counts were based on all days between the two average dates. This was repeated for all years from 1998 to 2022 when transect data were available. Bay segment stressor metrics were calculated as the average counts in each "transect year" from all stations in each segment. The stressor metrics were compared to frequency occurrence (all species) each year by bay segment. Generalized Linear regression Models (GLMs) were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, both, bay segment, and time period (recovery pre-2016, decline post-2016). Two models were evaluated, one with the temperature and salinity metrics together and another with the both metric given that the latter was a combined metric of the former. Interactions were included between temperature, salinity, and time period or the both metric and time period, depending on the model. Bay segment was included as a main effect without interactions. The most parsimonious models were identified by forward and backward selection of the independent variables and comparison with AIC values [@sakamoto1986; @venables2002b]. All models excluded Lower Tampa Bay because of minimal seagrass change over time.

@@

Long-term meteorological data showed increasing trends for air temperature and precipitation ([@fig-meteowqraw]). Mean annual air temperature has increased by `r airtempslo` $^\circ$C per year (`r airtemppvl`, $R^2$ `r airtemprsq`). Mean annual air temperature in 1975 was `r airtempstr` $^\circ$C, whereas current mean annual air temperature in 2022 was `r airtempend` $^\circ$C, showing an overall increase in the period of record of `r airtempdif` $^\circ$C. Similarly, total precipitation during the rainy season has increased by `r prpslo` mm per year, although the model was not significant (`r prppvl`, $R^2$ `r prprsq`). Removing September from the rainy season showed a significantly increasing trend of `r prpslo2` mm per year (`r prppvl2`, $R^2$ = `r prprsq2`). Using this model, total precipitation in 1975 was `r prpstr2` mm, whereas current total precipitation in 2022 was `r prpend2` mm, showing an overall increase in the period of record of `r prpdif2` mm. Trends in total annual hydrologic load from 1985 to present followed total annual precipitation, although the linear model was not significant likely due to a shorter period of record. The SPI showed notable anomalies in precipitation, with pronounced rainy periods in the early 1980s, late 1990s, 2005, and 2015-2020.
